
---
title: "Social support and domestic violence: ordinal logistic regression"
date: "March 20, 2026"
format: html
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

## Introduction

This report presents an ordinal logistic regression analysis using data collected via Prolific. The outcome variable is `ss1`, which captures participants' responses to the statement: *"I have someone who will listen to me when I need to talk."* Responses were recorded on a 5-point ordered scale from *Never* to *Always*.

The two explanatory variables are:

- **`nisvstot`**: Sum-score of a 34-item domestic violence questionnaire (yes/no items); higher scores indicate greater experience of violence and/or sexual assault.
- **`TBIcount`**: Total number of head injuries the participant has experienced.

The dataset contains `r nrow(pro2)` participants and `r ncol(pro2)` variables.

## Data loading and cleaning

```{r}
#| label: load-packages
library(tidyverse)
library(MASS)
library(nnet)
library(patchwork)
```

```{r}
#| label: load-data
load("/Users/homer/git/agnes/data/prolific.Rda")
```

```{r}
#| label: clean-data
ss_levels <- c("Never", "Rarely", "Sometimes", "Usually", "Always")

pro2 <- prolific |>
  mutate(
    s_english = factor(s_english, levels = c(0, 1), labels = c("No", "Yes"), ordered = TRUE),
    s_abuse   = factor(s_abuse,   levels = c(0, 1), labels = c("No", "Yes"), ordered = TRUE),
    across(c(ss1, ss2, ss3, ss4),
           ~ factor(.x, levels = 1:5, labels = ss_levels, ordered = TRUE))
  )
```

The following variables were transformed into ordered factors:

- `s_english` and `s_abuse`: binary No/Yes factors
- `ss1` through `ss4`: 5-point ordered factors (*Never* < *Rarely* < *Sometimes* < *Usually* < *Always*)

A note on missing data: `nisvstot` has `r sum(is.na(pro2$nisvstot))` missing values out of `r nrow(pro2)` rows (~`r round(mean(is.na(pro2$nisvstot)) * 100)`%), which reduces the effective sample size for the regression model.

## Ordinal logistic regression

An ordinal logistic regression model was fit using `MASS::polr()`, with `ss1` as the response variable and `nisvstot` and `TBIcount` as predictors.

```{r}
#| label: fit-model
fit_ord <- polr(ss1 ~ nisvstot + TBIcount, data = pro2, Hess = TRUE)
```

### Coefficients with p-values

P-values were computed using the normal approximation of the t-statistic.

```{r}
#| label: coef-pvalues
coef_table <- coef(summary(fit_ord))
p_values <- pnorm(abs(coef_table[, "t value"]), lower.tail = FALSE) * 2
coef_table <- cbind(coef_table, `p value` = p_values)
printCoefmat(coef_table, digits = 4, has.Pvalue = TRUE)
```

### Odds ratios with 95% confidence intervals

```{r}
#| label: odds-ratios
exp(cbind(OR = coef(fit_ord), confint(fit_ord)))
```

The odds ratio for `nisvstot` is below 1, indicating that higher domestic violence exposure is associated with lower odds of reporting higher perceived social support. The confidence interval for `TBIcount` includes 1, consistent with a non-significant result.

## Assumption checks

### Proportional odds assumption

The proportional odds assumption was tested by comparing the ordinal logistic regression model against a more flexible multinomial logistic regression model using a likelihood ratio test. A non-significant result supports the assumption.

```{r}
#| label: proportional-odds-test
fit_multi <- multinom(ss1 ~ nisvstot + TBIcount, data = pro2, trace = FALSE)

lrt_stat <- 2 * (logLik(fit_multi) - logLik(fit_ord))
df_diff  <- length(coef(fit_multi)) - length(c(coef(fit_ord), fit_ord$zeta))
p_lrt    <- pchisq(lrt_stat, df = df_diff, lower.tail = FALSE)

cat(sprintf("Chi-sq = %.3f, df = %d, p = %.4f\n", lrt_stat, df_diff, p_lrt))
```

The test was non-significant (p = `r round(p_lrt, 3)`), supporting the proportional odds assumption.

### Multicollinearity

```{r}
#| label: predictor-correlation
cor_val <- with(pro2, cor(nisvstot, TBIcount, use = "complete.obs"))
cat(sprintf("Pearson correlation between nisvstot and TBIcount: r = %.3f\n", cor_val))
```

The correlation between the two predictors is moderate (r ≈ `r round(with(pro2, cor(nisvstot, TBIcount, use = "complete.obs")), 2)`), which is not severe enough to substantially distort the estimates but is worth noting.

### Predictor distributions and relationship with ss1

```{r}
#| label: assumption-plots
#| fig-width: 8
#| fig-height: 6
model_data <- pro2 |>
  select(ss1, nisvstot, TBIcount) |>
  drop_na()

p1 <- ggplot(model_data, aes(x = nisvstot)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  labs(title = "Distribution of nisvstot", x = "nisvstot", y = "Count")

p2 <- ggplot(model_data, aes(x = TBIcount)) +
  geom_histogram(bins = 20, fill = "coral", color = "white") +
  labs(title = "Distribution of TBIcount", x = "TBIcount", y = "Count")

p3 <- ggplot(model_data, aes(x = nisvstot, y = as.numeric(ss1))) +
  geom_jitter(height = 0.2, alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  scale_y_continuous(breaks = 1:5, labels = levels(pro2$ss1)) +
  labs(title = "ss1 vs nisvstot", x = "nisvstot", y = "ss1")

p4 <- ggplot(model_data, aes(x = TBIcount, y = as.numeric(ss1))) +
  geom_jitter(height = 0.2, alpha = 0.4, color = "coral") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  scale_y_continuous(breaks = 1:5, labels = levels(pro2$ss1)) +
  labs(title = "ss1 vs TBIcount", x = "TBIcount", y = "ss1")

(p1 + p2) / (p3 + p4)
```

`TBIcount` shows notable right skew with a potential extreme outlier. This may be influencing the regression estimate for that predictor.

## Interpretation

### Threshold coefficients

The model estimates four threshold (intercept) parameters that define the boundaries between adjacent response categories. The `Never|Rarely` threshold of `r round(fit_ord$zeta["Never|Rarely"], 2)` is large and negative, reflecting the fact that very few participants (n = 4) reported *never* having someone to listen to them. When `nisvstot = 0` and `TBIcount = 0`, the predicted probability of responding *"Never"* is approximately 1 / (1 + exp(4.36)) ≈ 1.3% — consistent with high baseline perceived social support in this sample.

The `Usually|Always` threshold (≈ −0.015) is very close to zero, indicating that even at average predictor values, participants were roughly equally likely to respond *"Usually"* versus *"Always"* — reflecting the concentration of responses in the upper two categories.

### nisvstot

The coefficient for `nisvstot` is `r round(coef(fit_ord)["nisvstot"], 3)` (OR = `r round(exp(coef(fit_ord)["nisvstot"]), 3)`, p = 0.008). Each additional endorsement on the domestic violence questionnaire is associated with approximately a **7% decrease in the odds** of reporting a higher level of perceived social support. Across the observed range of scores (1–31), this represents a substantively meaningful effect: someone who endorsed 10 more items would have roughly half the odds of reporting higher social support compared to someone with 10 fewer endorsements.

This finding is clinically and theoretically interpretable: survivors of domestic violence and sexual assault may have had their support networks deliberately undermined by abusers, or may feel shame or fear that prevents them from reaching out.

### TBIcount

The coefficient for `TBIcount` is `r round(coef(fit_ord)["TBIcount"], 3)` (OR = `r round(exp(coef(fit_ord)["TBIcount"]), 3)`, p = 0.316). This effect is not statistically significant, and the 95% confidence interval for the odds ratio (0.94–1.22) includes 1. The number of head injuries does not appear to have a meaningful independent relationship with perceived social support in this sample, though the skewed distribution and potential extreme outlier warrant caution in interpreting this result.
