
---
title: "Social support and domestic violence: ordinal logistic regression"
date: "March 20, 2026"
format: html
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

## Introduction

This report presents an ordinal logistic regression analysis using data collected via Prolific. The outcome variable is `ss1`, which captures participants' responses to the statement: *"I have someone who will listen to me when I need to talk."* Responses were recorded on a 5-point ordered scale from *Never* to *Always*.

The two explanatory variables are:

- **`nisvstot`**: Sum-score of a 34-item domestic violence questionnaire (yes/no items); higher scores indicate greater experience of violence and/or sexual assault.
- **`TBIcount`**: Total number of head injuries the participant has experienced.

The dataset contains `r nrow(pro2)` participants and `r ncol(pro2)` variables.

## Data loading and cleaning

```{r}
#| label: load-packages
library(tidyverse)
library(MASS) # will mask tidyverse select
```

```{r}
#| label: load-data
load("/Users/homer/git/agnes/data/prolific.Rda")
```

```{r}
#| label: clean-data
ss_levels <- c("Never", "Rarely", "Sometimes", "Usually", "Always")

pro2 <- prolific |> 
  dplyr::select(ss1, TBIcount) |> 
  mutate(ss1_2 = haven::as_factor(ss1)) |> 
  mutate(ss1_2 = factor(ss1_2, levels = ss_levels))
```



## Bootstrap Example

```{r}
library(rsample)
library(purrr)
library(broom) # will mask rsample tidy

set.seed(8214)

# Use the same complete-case data as the ordinal model
model_data <- pro2 |>
  dplyr::select(ss1_2, TBIcount) |>
  drop_na()

boot_splits <- bootstraps(model_data, times = 1000)

# Fit on each bootstrap resample
fit_on_split <- function(split) {
  MASS::polr(ss1_2 ~ TBIcount,
      data = rsample::analysis(split),
      Hess = TRUE)
}

summary(fit_on_split(boot_splits$splits[[2]]))

boot_results <- boot_splits |>
  mutate(
    model = map(splits, fit_on_split),
    coefs = map(model, broom::tidy)
  )

# Unnest and compute percentile CIs
boot_ci <- boot_results |>
  dplyr::select(id, coefs) |>
  unnest(coefs) |>
  summarise(
    boot_estimate = mean(estimate),
    lower_95 = quantile(estimate, 0.025),
    upper_95 = quantile(estimate, 0.975),
    .by = term
  )

boot_ci
```

You would still want to check the proportional odds assumption and look at correlations between [redictors (see other Quarto document).

